---
title: "Adv. ML. Lab 2"
author: "Simon Jorstedt"
date: "`r Sys.Date()`"
output: pdf_document
---

The game.


```{r}

```




# Problem 1


We will assume that the robot starts in a random state.


```{r}
# This chunk is for like, loading packages, and like doing other kinds of setup
# like creating the starting probabilities and the transition probability matrix.

library(HMM)

startProbs <- rep(0.1, 10)

transProbs <- matrix(c(c(0.5, 0.5,  0,   0,   0,   0,   0,   0,   0,     0),
                       c(0,   0.5,  0.5, 0,   0,   0,   0,   0,   0,     0),
                       c(0,   0,    0.5, 0.5, 0,   0,   0,   0,   0,     0),
                       c(0,   0,    0,   0.5, 0.5, 0,   0,   0,   0,     0),
                       c(0,   0,    0,   0,   0.5, 0.5, 0,   0,   0,     0),
                       c(0,   0,    0,   0,   0,   0.5, 0.5, 0,   0,     0),
                       c(0,   0,    0,   0,   0,   0,   0.5, 0.5, 0,     0),
                       c(0,   0,    0,   0,   0,   0,   0,   0.5, 0.5,   0),
                       c(0,   0,    0,   0,   0,   0,   0,   0,   0.5,   0.5),
                       c(0.5, 0,    0,   0,   0,   0,   0,   0,   0,     0.5)
                       ), nrow = 10, ncol = 10, byrow = TRUE)

emissionProbs <- matrix(c(c(0.2, 0.2, 0.2, 0.0, 0.0, 0.0, 0.0, 0.0, 0.2, 0.2),
                          c(0.2, 0.2, 0.2, 0.2, 0.0, 0.0, 0.0, 0.0, 0.0, 0.2),
                          c(0.2, 0.2, 0.2, 0.2, 0.2, 0.0, 0.0, 0.0, 0.0, 0.0),
                          c(0.0, 0.2, 0.2, 0.2, 0.2, 0.2, 0.0, 0.0, 0.0, 0.0),
                          c(0.0, 0.0, 0.2, 0.2, 0.2, 0.2, 0.2, 0.0, 0.0, 0.0),
                          c(0.0, 0.0, 0.0, 0.2, 0.2, 0.2, 0.2, 0.2, 0.0, 0.0),
                          c(0.0, 0.0, 0.0, 0.0, 0.2, 0.2, 0.2, 0.2, 0.2, 0.0),
                          c(0.0, 0.0, 0.0, 0.0, 0.0, 0.2, 0.2, 0.2, 0.2, 0.2),
                          c(0.2, 0.0, 0.0, 0.0, 0.0, 0.0, 0.2, 0.2, 0.2, 0.2),
                          c(0.2, 0.2, 0.0, 0.0, 0.0, 0.0, 0.0, 0.2, 0.2, 0.2)
                          ), nrow = 10, ncol = 10, byrow = TRUE)

# Initialise a HMM model
HMM_model <- initHMM(States = 1:10,
                     Symbols = 1:10,
                     startProbs = startProbs,
                     transProbs = transProbs,
                     emissionProbs = emissionProbs)
```

# Problem 2
Now we shall simulate the HMM that we initiated previously. This comes down to just running the `simHMM` function.

```{r}
set.seed(322347)
HMM_run1 <- simHMM(HMM_model, length = 100)
```

# Problem 3
Now we shall use our emission observations above to compute the smoothed and filtered probability distributions respectively using the Forward-Backward algorithm. We will also compute the most probable path using the Viterbi algorithm.




